# SBA loan data analysis
## 1. Setting up and loading/cleaning data
### 1.1 Load libraries, mappings, and data
```{r, results = 'hide', warning = FALSE, message=FALSE}
library(tidyverse)
library(geomtextpath)
library(plotly)
library(htmlwidgets)
library(maps)
library(dplyr)
df <- read.csv("SBAnational.csv")
naics_mapping <- list(
  "11" = "Agriculture, Forestry, Fishing and Hunting",
  "21" = "Mining, Quarrying, and Oil and Gas Extraction",
  "22" = "Utilities",
  "23" = "Construction",
  "31" = "Manufacturing",
  "32" = "Manufacturing",
  "33" = "Manufacturing",
  "42" = "Wholesale Trade",
  "44" = "Retail Trade",
  "45" = "Retail Trade",
  "48" = "Transportation and Warehousing",
  "49" = "Transportation and Warehousing",
  "51" = "Information",
  "52" = "Finance and Insurance",
  "53" = "Real Estate and Rental and Leasing",
  "54" = "Professional, Scientific, and Technical Services",
  "55" = "Management of Companies and Enterprises",
  "56" = "Administrative and Support Services",
  "61" = "Educational Services",
  "62" = "Health Care and Social Assistance",
  "71" = "Arts, Entertainment, and Recreation",
  "72" = "Accommodation and Food Services",
  "81" = "Other Services",
  "92" = "Public Administration"
)
```
### 1.2 Filter out missing variables and limit data range to 2000-2010
```{r}
df <- df %>%
  filter(df$MIS_Status != "", df$GrAppv != "", df$SBA_Appv != "",
         df$NAICS != 0, df$State != "", df$NoEmp <= 1500,
         df$LowDoc == "Y" | df$LowDoc == "N", df$Term > 0,
         df$ApprovalFY < 2011 & df$ApprovalFY > 1999, df$NewExist != 0,
         df$RevLineCr == "Y" | df$RevLineCr == "N", df$UrbanRural > 0)

df$GrAppv <- as.numeric(gsub("[$,]", "", df$GrAppv))
df$SBA_Appv <- as.numeric(gsub("[$,]", "", df$SBA_Appv))
df$DisbursementGross <- as.numeric(gsub("[$,]", "", df$DisbursementGross))
df$ApprovalFY <- as.numeric(df$ApprovalFY)
```
### 1.3 Convert variables as needed and select desired variables
```{r}
df <- df %>%
  mutate(RealEstate = ifelse(df$Term >= 240, 1, 0)) %>%
  mutate(MIS_Status = ifelse(df$MIS_Status == "CHGOFF", 1, 0)) %>%
  mutate(FranchiseCode = ifelse(df$FranchiseCode > 1, 1, 0)) %>%
  mutate(UrbanRural = ifelse(df$UrbanRural == 1, 1, 0)) %>%
  mutate(NewExist = ifelse(df$NewExist == 1, 1, 0)) %>%
  mutate(RevLineCr = ifelse(df$RevLineCr == "Y", 1, 0)) %>%
  mutate(LowDoc = ifelse(df$LowDoc == "Y", 1, 0)) %>%
  mutate(NAICS = as.integer(NAICS / 10000)) %>%
  mutate(Industry = as.character(naics_mapping[as.character(NAICS)])) %>%
  mutate(Proportion = SBA_Appv / GrAppv) %>%
  mutate(Recession = ifelse(df$ApprovalFY > 2007 & df$ApprovalFY <= 2009,
                            1, 0))
df <- df %>%
  select(State, Industry, FranchiseCode, UrbanRural,
         Term, NoEmp, NewExist, RevLineCr, Recession,
         RealEstate, DisbursementGross, LowDoc, NoEmp, Proportion, MIS_Status)
```
### 1.4 Recession anaylsis
Determine whether or not data that occurred in the recession is skewed enough to warrant excluding it from the data set. Consider p-value < 0.05 significant
```{r}
recession <- df %>%
  select(MIS_Status, Recession)
chisq.test(table(recession))
```
```{r}
recession_summary <- recession %>%
  group_by(Recession, MIS_Status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Recession) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

recession_plot <- ggplotly(
  ggplot(recession_summary, aes(
    x = factor(Recession),
    y = percentage,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
    text = paste(
      "Loan Occured during the Recession: ", ifelse(Recession == 1, "Yes",
                                                    "No"),
      "<br>Default Status: ", ifelse(MIS_Status == 1, "Defaulted",
                                     "Paid In Full"),
      "<br>Percentage: ", scales::percent(percentage, accuracy = 0.01)
    )
  )) +
    geom_bar(stat = "identity", position = "fill", alpha = 0.75) +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    labs(
      title = "Loan Occured during the Recession vs Default Rate",
      x = "Loan Occured during the Recession",
      y = "Percentage",
      fill = "Default Status"
    ),
  tooltip = "text"
)

recession_plot
```
With a p-value <0.05 and there being a significant difference in default rate between loans that occurred during the Great Recession and loans that didn't occur during the Great Recession we will exclude the data from the Great Recession

```{r}
df <- df %>%
  filter(df$Recession != 1)
df <- df %>% select(-Recession)
```

## 2. Statistcal Analysis
### 2.1 State
```{r}
state_mapping <- setNames(state.name, state.abb)
state_test <- df %>%
  select(State, MIS_Status)
chisq.test(table(state_test))
```
```{r}
state <- df %>%
  mutate(StateName = tolower(state_mapping[State])) %>%
  group_by(StateName) %>%
  summarise(default_rate = mean(MIS_Status == 1))
map <- map_data("state")
map <- left_join(map, state, by = c("region" = "StateName"))
state_plot <- ggplotly(
  ggplot(map, aes(x = long, y = lat, group = group, fill = default_rate,
    text = paste(
      "State: ", tools::toTitleCase(region),
      "<br>Default Rate: ", scales::percent(default_rate, accuracy = 0.01)
    )
  )) +
    geom_polygon(color = "white") +
    scale_fill_gradient(low = "lightblue", high = "red",
                        labels = scales::percent) +
    labs(
      title = "Loan Default Rate by State",
      fill = "Default Rate (%)"
    ),
  tooltip = "text"
)
state_plot
```
## 2.2 Industry
```{r}
industry <- df %>%
  select(MIS_Status, Industry)
chisq.test(table(industry))
```
Create two summaries, one that includes both default and paid in full rates for graphing, one that includes only default rate for later use.
```{r}
industry_summary <- industry %>%
  group_by(Industry, MIS_Status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Industry) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

industry_summary2 <- industry %>%
  group_by(Industry) %>%
  summarise(default_rate = mean(MIS_Status == 1))
```
```{r}
industry_plot <- ggplotly(
  ggplot(industry_summary, aes(
    x = factor(Industry),
    y = percentage,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
    text = paste(
      "Industry: ", Industry,
      "<br>Default Status: ", ifelse(MIS_Status == 1, "Defaulted",
                                     "Paid In Full"),
      "<br>Percentage: ", scales::percent(percentage, accuracy = 0.01)
    )
  )) +
    geom_bar(stat = "identity", position = "fill", alpha = 0.75) +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
      title = "Industry vs Default Rate",
      x = "Industry",
      y = "Percentage",
      fill = "Default Status"
    ),
  tooltip = "text"
)
industry_plot
```
## 2.3 Real Estate
```{r}
real_estate <- df %>%
  select(MIS_Status, RealEstate)
chisq.test(table(real_estate))
```
```{r}
real_estate_summary <- real_estate %>%
  group_by(RealEstate, MIS_Status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(RealEstate) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

real_estate_plot <- ggplotly(
  ggplot(real_estate_summary, aes(
    x = factor(RealEstate),
    y = percentage,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
    text = paste(
      "Backed by Real Estate: ", ifelse(RealEstate == 1, "Yes", "No"),
      "<br>Default Status: ", ifelse(MIS_Status == 1, "Defaulted",
                                     "Paid In Full"),
      "<br>Percentage: ", scales::percent(percentage, accuracy = 0.01)
    )
  )) +
    geom_bar(stat = "identity", position = "fill", alpha = 0.75) +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    labs(
      title = "Real Estate vs Default Rate",
      x = "Backed By Real Estate",
      y = "Percentage",
      fill = "Default Status"
    ),
  tooltip = "text"
)

real_estate_plot
```
### 2.4 Franchise
```{r}
franchise <- df %>%
  select(MIS_Status, FranchiseCode)
chisq.test(table(franchise))
```
```{r}
franchise_summary <- franchise %>%
  group_by(FranchiseCode, MIS_Status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(FranchiseCode) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

franchise_plot <- ggplotly(
  ggplot(franchise_summary, aes(
    x = factor(FranchiseCode),
    y = percentage,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
    text = paste(
      "Is Franchise: ", ifelse(FranchiseCode == 1, "Yes", "No"),
      "<br>Default Status: ", ifelse(MIS_Status == 1, "Defaulted",
                                     "Paid In Full"),
      "<br>Percentage: ", scales::percent(percentage, accuracy = 0.01)
    )
  )) +
    geom_bar(stat = "identity", position = "fill", alpha = 0.75) +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    labs(
      title = "Is Franchise vs Default Rate",
      x = "Is Franchise",
      y = "Percentage",
      fill = "Default Status"
    ),
  tooltip = "text"
)

franchise_plot
```
### 2.5 New vs Existing businesses
```{r}
new <- df %>%
  select(MIS_Status, NewExist)
chisq.test(table(new))
```
```{r}
new_summary <- new %>%
  group_by(NewExist, MIS_Status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(NewExist) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

new_plot <- ggplotly(
  ggplot(new_summary, aes(
    x = factor(NewExist),
    y = percentage,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
    text = paste(
      "Existed for more than 2 years: ", ifelse(NewExist == 1, "Yes", "No"),
      "<br>Default Status: ", ifelse(MIS_Status == 1, "Defaulted",
                                     "Paid In Full"),
      "<br>Percentage: ", scales::percent(percentage, accuracy = 0.01)
    )
  )) +
    geom_bar(stat = "identity", position = "fill", alpha = 0.75) +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    labs(
      title = "Is Existing vs Default Rate",
      x = "Is Existing",
      y = "Percentage",
      fill = "Default Status"
    ),
  tooltip = "text"
)

new_plot
```
### 2.6 Revolving Line of Credit
```{r}
credit <- df %>%
  select(MIS_Status, RevLineCr)
chisq.test(table(credit))

```
```{r}
credit_summary <- credit %>%
  group_by(RevLineCr, MIS_Status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(RevLineCr) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

credit_plot <- ggplotly(
  ggplot(credit_summary, aes(
    x = factor(RevLineCr),
    y = percentage,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
    text = paste(
      "Has Revolving Line of Credit: ", ifelse(RevLineCr == 1, "Yes", "No"),
      "<br>Default Status: ", ifelse(MIS_Status == 1, "Defaulted",
                                     "Paid In Full"),
      "<br>Percentage: ", scales::percent(percentage, accuracy = 0.01)
    )
  )) +
    geom_bar(stat = "identity", position = "fill", alpha = 0.75) +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    labs(
      title = "Has Revolving Line of Credit vs Default Rate",
      x = "Has Revolving Line of Credit",
      y = "Percentage",
      fill = "Default Status"
    ),
  tooltip = "text"
)

credit_plot
```
### 2.7 Urban vs Rural
```{r}
urban <- df %>%
  select(MIS_Status, UrbanRural)
chisq.test(table(urban))
```
```{r}
urban_summary <- urban %>%
  group_by(UrbanRural, MIS_Status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(UrbanRural) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

urban_plot <- ggplotly(
  ggplot(urban_summary, aes(
    x = factor(UrbanRural),
    y = percentage,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
    text = paste(
      "Urban or Rural: ", ifelse(UrbanRural == 1, "Urban", "Rural"),
      "<br>Default Status: ", ifelse(MIS_Status == 1, "Defaulted",
                                     "Paid In Full"),
      "<br>Percentage: ", scales::percent(percentage, accuracy = 0.01)
    )
  )) +
    geom_bar(stat = "identity", position = "fill", alpha = 0.75) +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "Rural", "1" = "Urban")) +
    labs(
      title = "Ubran/Rural vs Default Rate",
      x = "Ubran/Rural",
      y = "Percentage",
      fill = "Default Status"
    ),
  tooltip = "text"
)

urban_plot
```
### 2.8 LowDoc loan program
```{r}
lowdoc <- df %>%
  select(MIS_Status, LowDoc)
chisq.test(table(lowdoc))
```
```{r}
lowdoc_summary <- lowdoc %>%
  group_by(LowDoc, MIS_Status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(LowDoc) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()

lowdoc_plot <- ggplotly(
  ggplot(lowdoc_summary, aes(
    x = factor(LowDoc),
    y = percentage,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
    text = paste(
      "In LowDoc program: ", ifelse(LowDoc == 1, "Yes", "No"),
      "<br>Default Status: ", ifelse(MIS_Status == 1, "Defaulted",
                                     "Paid In Full"),
      "<br>Percentage: ", scales::percent(percentage, accuracy = 0.01)
    )
  )) +
    geom_bar(stat = "identity", position = "fill", alpha = 0.75) +
    scale_y_continuous(labels = scales::percent) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    labs(
      title = "In LowDoc Program vs Default Rate",
      x = "In LowDoc Programt",
      y = "Percentage",
      fill = "Default Status"
    ),
  tooltip = "text"
)

lowdoc_plot
```
### 2.9 SBA gauranteed proportion
```{r}
sba_prop <- df %>%
  select(MIS_Status, Proportion)
t.test(Proportion ~ MIS_Status, data = sba_prop, alternative = "two.sided")
```
```{r}
sba_prop_plot <- ggplotly(
  ggplot(sba_prop, aes(
    x = Proportion,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
                   text = paste(
                     "Term: ", after_stat(xmin), " to ", after_stat(xmax),
                     "<br>Count: ", after_stat(count)
                   ))) +
    geom_histogram(
                   position = "dodge",
                   binwidth = 0.25,
                   boundary = 0,
                   alpha = 0.7) +
    xlim(0, max(sba_prop$Proportion)) +
    facet_wrap(~ MIS_Status,
               scales = "free_y",
               labeller = as_labeller(c("0" = "Paid In Full",
                                        "1" = "Defaulted"))) +
    labs(
      title = "Histogram of loan portion gauranteed by MIS Status",
      x = "Porportion of loan gauranteed by SBA",
      y = "Count",
      fill = "Default Status"
    ),
  tooltip = "text"
)
sba_prop_plot
```
### 2.10 Gross Disbursement
Find quartile for data to make a barchart that displays the distribution of gross disbursement by quartile 
```{r}
disbur <- df %>%
  mutate(
    Quartile = cut(
      DisbursementGross,
      breaks = quantile(DisbursementGross, probs = seq(0, 1, by = 0.25)),
      include.lowest = TRUE,
      labels = c("0-25%", "25-50%", "50-75%", "75-100%")
    )
  ) %>%
  select(MIS_Status, DisbursementGross, Quartile)
t.test(DisbursementGross ~ MIS_Status, data = disbur,
       alternative = "two.sided")
```
```{r}
disbur_prop <- disbur %>%
  group_by(MIS_Status, Quartile) %>%
  summarise(Count = n()) %>%
  group_by(MIS_Status) %>%
  mutate(Proportion = Count / sum(Count))
disbur_plot <- ggplotly(
  ggplot(disbur_prop, aes(x = Quartile, y = Proportion, fill = Quartile,
   text = paste(
    "Quartile: ", Quartile,
    "<br>Percentage: ", scales::percent(Proportion, accuracy = 0.01)
  ))) +
    geom_bar(stat = "identity", alpha = 0.75) +
    facet_wrap(~ MIS_Status, labeller =
                 as_labeller(c(`0` = "Paid In Full", `1` = "Default"))) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = "Gross Disbursement Distribution by Quartiles",
      x = "Gross Disbursement Quartile",
      y = "Proportion",
      fill = "Quartile"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5)
    ),
  tooltip = "text"
)
disbur_plot
```
### 2.11 Number of employees
Do same process as done with the gross disbursement
```{r}
emp <- df %>%
  mutate(
    Quartile = cut(
      NoEmp,
      breaks = quantile(NoEmp, probs = seq(0, 1, by = 0.25)),
      include.lowest = TRUE,
      labels = c("0-25%", "25-50%", "50-75%", "75-100%")
    )
  ) %>%
  select(MIS_Status, NoEmp, Quartile)
t.test(NoEmp ~ MIS_Status, data = emp,
       alternative = "two.sided")
```
```{r}
emp_prop <- emp %>%
  group_by(MIS_Status, Quartile) %>%
  summarise(Count = n()) %>%
  group_by(MIS_Status) %>%
  mutate(Proportion = Count / sum(Count))
emp_plot <- ggplotly(
  ggplot(emp_prop, aes(x = Quartile, y = Proportion, fill = Quartile,
  text = paste (
    "Quartile: ", Quartile,
    "<br>Percentage: ", scales::percent(Proportion, accuracy = 0.01)
  ))) +
    geom_bar(stat = "identity", alpha = 0.75) +
    facet_wrap(~ MIS_Status, labeller =
                 as_labeller(c(`0` = "Paid In Full", `1` = "Default"))) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = "Number of employee Distribution by Quartiles",
      x = "Disbursement Quartile",
      y = "Proportion",
      fill = "Quartile"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      strip.text = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5)
    ),
  tooltip = "text"
)
emp_plot
```
### 2.13 Term Length
```{r}
term <- df %>%
  select(MIS_Status, Term)
t.test(Term ~ MIS_Status, data = term, alternative = "two.sided")
```
Filter few data points to make graph more balanced
```{r}
term <- term %>%
  filter(Term <= 300)
```
```{r}
term_plot <- ggplotly(
  ggplot(term, aes(
    x = Term,
    fill = factor(MIS_Status, labels = c("Paid In Full", "Defaulted")),
                   text = paste(
                     "Term: ", after_stat(xmin), " to ", after_stat(xmax),
                     "<br>Count: ", after_stat(count)
                   ))) +
    geom_histogram(
                   position = "dodge",
                   binwidth = 25,
                   boundary = 0,
                   alpha = 0.7) +
    xlim(0, max(term$Term)) +
    facet_wrap(~ MIS_Status,
               scales = "free_y",
               labeller = as_labeller(c("0" = "Paid In Full",
                                        "1" = "Defaulted"))) +
    labs(
      title = "Histogram of Term by MIS Status",
      x = "Term",
      y = "Count",
      fill = "Default Status"
    ),
  tooltip = "text"
)
term_plot
```



















